---
title: "RNA Seq"
output: html_notebook
---

# First Load All Libraries
```{r}
library(limma)
library(DESeq2)
library(ggplot2)
library(tidyverse)
library(readr)
library(ggrepel)
library(clusterProfiler)
library(org.Mm.eg.db)   # change to org.Hs.eg.db if human
library(enrichplot)
library(DOSE)
library(readxl)
library(AnnotationDbi)
library(stringr)
library(dplyr)
library(pheatmap)
library(umap)
library(plotly)
library(fgsea)
library(caret)
library(glmnet)
library(xgboost)
```


```{r}
FastP_featurecounts <- read.csv("FastP_featurecounts.csv")
metadata <- read.csv("metadata.csv")
```

# Cleaned up Data Matrix
```{r}
head(FastP_featurecounts)

fc <- FastP_featurecounts[,-(2:6)]
colnames(fc) <- as.character(unlist(fc[1,]))
fc <- fc[-1,]
rownames(fc) <- as.character(unlist(fc[,1]))
fc <- fc[,-1]
head(fc)

count_matrix <- as.matrix(fc)
head(count_matrix)
storage.mode(count_matrix) <- "numeric"
print(is.numeric(count_matrix))
```

Adding Column for Ensembl IDs to Gene IDs
```{r}
ids <- data.frame(ensembl_ids = rownames(fc), stringsAsFactors = FALSE)

# 2. Clean Ensembl IDs (remove .version)
ids$clean_ensembl <- str_remove(ids$ensembl_ids, "\\.\\d+$")

# 3. Query annotations
annotations <- AnnotationDbi::select(org.Mm.eg.db,
                      keys = unique(ids$clean_ensembl),
                      columns = c("SYMBOL", "GENENAME", "ENTREZID"),
                      keytype = "ENSEMBL")

# 4. Enforce one-to-one mapping
annot_unique <- annotations[!duplicated(annotations$ENSEMBL), ]
rownames(annot_unique) <- annot_unique$ENSEMBL

# 5. Build master table by matching CLEAN IDs
ids$SYMBOL   <- annot_unique[ids$clean_ensembl, "SYMBOL"]
ids$GENENAME <- annot_unique[ids$clean_ensembl, "GENENAME"]
ids$ENTREZID <- annot_unique[ids$clean_ensembl, "ENTREZID"]

# 6. Inspect
head(ids)
```


# Check Files are Loaded in and begin cleaning of them
```{r}
# Check head of read in files, and then ccreate copies ot modify
head(metadata)
meta <- metadata

# Correct unique grouping so that we can quickly split on it
meta <- meta %>%
  mutate(
    GroupID = group_indices(., Age, Exposure, Diet, Treatment)
  )

meta %>%
  distinct(Age, Exposure, Diet, Treatment, GroupID) %>%
  arrange(GroupID)

meta$SampleID <- rownames(meta)


meta <- meta %>%
  mutate(
    plot_group = paste0(Exposure,Diet,Treatment, " ", Age)
  )
```
Now we want to compute basic DESeq2 Object and Limma Object, with transforms of scale and quantile.
```{r}
# Run DESeq2 Object along with most result important genes, along with variance transform
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData   = metadata,
  design    = ~ Age + Exposure + Diet + Treatment
)

#Potential Filtering Option
dds <- dds[rowSums(counts(dds)) >= 10, ]

dds <- DESeq(dds)

res <- results(dds)

res <- res[order(res$padj),]
head(res)

vsd <- vst(dds, blind=FALSE)
vst <- assay(vsd)
```
Exporting VST to CSV
```{r}
vst_df <- as.data.frame(vst)
head(vst_df)

#write.csv(vsd, "DESeq_VST.csv",row.names = TRUE)
#write.csv(vst_filt, "DESeq_Filtered_VST.csv", row.names = TRUE)
```


Limma Array Initializing
```{r}
log_counts <- log2(count_matrix+1)

qn_log_counts <- normalizeBetweenArrays(
     log_counts,
     method = "quantile")
```

Cleaning the rownames of the Meta Dataframe
```{r}
meta <- as.data.frame(meta)
meta$SampleID <- sprintf("Yim-%03d", as.integer(rownames(meta)))
rownames(meta) <- meta$SampleID
meta$SampleID <- NULL

meta



```
We want to do Density Plots for the Raw Counts, DESeq2 VST Counts, and the QN_Log_Counts
```{r}
outliers <- c("Yim-009","Yim-015","Yim-058","Yim-048","Yim-071","Yim-006","Yim-017",
              "Yim-045","Yim-034","Yim-039")

plot_density_with_colored_outliers <- function(mat, title,
                                               log_transform = FALSE,
                                               xlim_zoom = NULL,
                                               ylim_zoom = NULL) {
  
  df <- as.data.frame(mat)
  df$Gene <- rownames(df)
  
  df_long <- df %>%
    pivot_longer(
      cols = -Gene,
      names_to = "Sample",
      values_to = "Value"
    ) %>%
    mutate(
      ColorGroup = ifelse(Sample %in% outliers, Sample, "Other"),
      PlotValue = if (log_transform) log2(Value + 1) else Value
    ) %>%
    arrange(ColorGroup == "Other")  # gray lines first, outliers on top

  p <- ggplot(df_long, aes(x = PlotValue, group = Sample, color = ColorGroup)) +
    geom_density(linewidth = 0.8) +  # lines only
    scale_color_manual(
      values = c(
        "Other" = "grey80",
        setNames(scales::hue_pal()(length(outliers)), outliers)
      )
    ) +
    # filled legend trick
    guides(color = guide_legend(
      override.aes = list(linetype = 0, shape = 22, fill = c("grey80", scales::hue_pal()(length(outliers))))
    )) +
    labs(
      title = title,
      x = if (log_transform) "log2(count + 1)" else "Expression",
      y = "Density",
      color = "Sample"
    ) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold"))
  
  if (!is.null(xlim_zoom) || !is.null(ylim_zoom)) {
    p <- p + coord_cartesian(xlim = xlim_zoom, ylim = ylim_zoom)
  }
  
  return(p)
}




plot_density_with_colored_outliers(
  mat = count_matrix,
  title = "Raw Counts Density",
  log_transform = TRUE,
  xlim_zoom = c(0, 3),
  ylim_zoom = c(0, 0.175)
)

plot_density_with_colored_outliers(
  mat = vsd,
  title = "VST Density (DESeq2 Variance Stabilized Transform)",
  log_transform = FALSE,
  xlim_zoom = c(5, 6),
  ylim_zoom = c(0, 0.6)
)

plot_density_with_colored_outliers(
  mat = qn_log_counts,
  title = "Quantile-Normalized Log Density",
  log_transform = FALSE,
  xlim_zoom = c(0, 3),
  ylim_zoom = c(0, 0.2)
)


```
Computing PCA for DESeq Filtered
```{r}

pca_input <- t(vst)

any(apply(pca_input, 2, var) == 0)


pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  metadata[rownames(pca$x), ]
)

pve <- pca$sdev^2 / sum(pca$sdev^2)
pve[1:5]


plot(
  pca_df$PC1, pca_df$PC2,
  xlab = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
  ylab = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
  pch = 19
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)

pca_df$GroupFactor <- meta[rownames(pca_df), "Group"]
#pca_df$GroupFactor <- factor(pca_df$GroupID)


ggplot(pca_df, aes(x = PC1, y = PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of 8 Groups",
    subtitle="DESeq Filtered Beforehand"
  ) +
  theme_classic()


```

Computing Saline PCA with DESeq Filtered
```{r}
keep<- !(
  rownames(meta) == "Yim-009" |
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
vst_sub <- vst[, rownames(meta_sub)]

pca_input <- t(vst_sub)
any(apply(pca_input, 2, var) == 0)
pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of DESeq Filtered Saline Data",
    subtitle="Sample 9 is removed for Clarity"
  ) +
  theme_classic()
```

Computing Saline PCA with Sample 9 in it with DESeq Filtered
```{r}
keep<- !(
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
vst_sub <- vst_filt[, rownames(meta_sub)]

pca_input <- t(vst_filt_sub)

pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of DESeq Filtered Saline Data",
    subtitle="Sample 9 is not removed"
  ) +
  theme_classic()
```


Computing PCA for Quantile Normalization
```{r}
dim(qn_log_counts)              # genes x samples
mode(qn_log_counts)             # should be "numeric"
anyNA(qn_log_counts)            # should be FALSE

gene_var <- apply(qn_log_counts, 1, var)

qn_log_filt <- qn_log_counts[gene_var > 1e-8, ]

pca_input <- t(qn_log_filt)

any(apply(pca_input, 2, var) == 0)


pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  metadata[rownames(pca$x), ]
)

pve <- pca$sdev^2 / sum(pca$sdev^2)
pve[1:5]


plot(
  pca_df$PC1, pca_df$PC2,
  xlab = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
  ylab = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
  pch = 19
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)

pca_df$GroupFactor <- meta[rownames(pca_df), "Group"]
#pca_df$GroupFactor <- factor(pca_df$GroupID)


ggplot(pca_df, aes(x = PC1, y = PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of 8 Groups",
    subtitle="Sample 9 Included along with Ages"
  ) +
  theme_classic()


```

PCA of Quantile with Sample 9 removed
```{r}
keep<- !(
  rownames(meta) == "Yim-009" |
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
qn_sub <- qn_log_counts[, rownames(meta_sub)]

gene_var <- apply(qn_sub, 1, var)
qn_sub_filt <- qn_sub[gene_var > 1e-8, ]

pca_input <- t(qn_sub_filt)

stopifnot(!any(apply(pca_input, 2, var) == 0))

pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of Quantile Normalized Saline Data",
    subtitle="Sample 9 is Removed for Clarity"
  ) +
  theme_classic()
```

Having PCA with sample 9 in it
```{r}
keep<- !(
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
qn_sub <- qn_log_counts[, rownames(meta_sub)]

gene_var <- apply(qn_sub, 1, var)
qn_sub_filt <- qn_sub[gene_var > 1e-8, ]

pca_input <- t(qn_sub_filt)

stopifnot(!any(apply(pca_input, 2, var) == 0))

pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of Quantile Normalized Saline Data",
    subtitle="Sample 9 is is not removed"
  ) +
  theme_classic()

```
Now we will extract these PCA loadings, and get the most important genes from them.
```{r}
loadings <- as.data.frame(pca$rotation[, 1:2])
colnames(loadings) <- c("PC1", "PC2")
loadings$GeneID <- rownames(loadings)

top_pc1 <- loadings |>
  dplyr::arrange(desc(abs(PC1))) |>
  dplyr::slice(1:20)

top_pc2 <- loadings |>
  dplyr::arrange(desc(abs(PC2))) |>
  dplyr::slice(1:20)

top_pc1$Direction <- ifelse(top_pc1$PC1 > 0, "Positive", "Negative")
top_pc2$Direction <- ifelse(top_pc2$PC2 > 0, "Positive", "Negative")

pc1_genes <- top_pc1$GeneID
pc2_genes <- top_pc2$GeneID

#cat(pc1_genes,sep="\n")
#print("DIVIDER\n")
#cat(pc2_genes,sep="\n")

pc1_gene_names <- c("Myh9",
"Rnf11",
"Cltc",
"Col4a1",
"Slc10a1",
"Iqgap1",
"Scrn2",
"Cfap68",
"Hsd11b1",
"Actr2",
"Agfg1",
"Mrps35",
"Sqstm1",
"Fat1",
"Ctnnb1",
"Aamdc",
"Flna",
"Nme3",
"BC024139",
"Ces3b")

pc2_gene_names <- c("Fhit",
"Arhgef16",
"Qprt",
"Hectd1",
"Mfsd12",
"Eno1",
"Dera",
"Acadm",
"Gm70036",
"Nudt7",
"Hsd17b6",
"Cndp2",
"Napa",
"Klk1b4",
"Tmem86b",
"Dnaja3",
"Pgd",
"Frmd8",
"Slc23a1",
"Pcbd1")

top_pc1$GeneName <- pc1_gene_names
top_pc2$GeneName <- pc2_gene_names

top_pc1$Magnitude <- abs(top_pc1$PC1)
top_pc2$Magnitude <- abs(top_pc2$PC2)

top_pc1$GeneName <- reorder(
  top_pc1$GeneName,
  top_pc1$Magnitude
)

top_pc2$GeneName <- reorder(
  top_pc2$GeneName,
  top_pc2$Magnitude
)

ggplot(top_pc1,
       aes(x = GeneName, y = Magnitude, fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "Positive" = "blue",
    "Negative" = "red"
  )) +
  labs(
    x = "Gene",
    y = "PC1 Loading",
    title = "Top Genes Driving PC1"
  ) +
  theme_classic()

ggplot(top_pc2,
       aes(x = GeneName, y = Magnitude, fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "Positive" = "blue",
    "Negative" = "red"
  )) +
  labs(
    x = "Gene",
    y = "PC2 Loading",
    title = "Top Genes Driving PC2"
  ) +
  theme_classic()

```

This is the point where we will now start comparing between ages and then the differences from these and groups. We will start with just using DESeq
```{r}
dds_ctrl <- dds[, dds$Group == "FAPDsa"]

design(dds_ctrl) <- ~ Age
dds_ctrl <- DESeq(dds_ctrl)

head(dds_ctrl)

res_ctrl_time <- results(dds_ctrl, contrast = c("Age", "1 day", "28 day"))
res_ctrl_time <- res_ctrl_time[order(res_ctrl_time$padj), ]
head(res_ctrl_time)

sig <- subset(res_ctrl_time, padj < 0.05)

up   <- sig[sig$log2FoldChange > 0, ]
down <- sig[sig$log2FoldChange < 0, ]

```

We will now create a function so that we can quickly run the DESeq Object with whatever new group we would like to test.
```{r}
run_age_comparison <- function(dds, control_group, test_group, age_col = "Age", group_col = "Group") {

  # -------------------------
  # Subset to the two groups
  # -------------------------
  keep <- dds[[group_col]] %in% c(control_group, test_group)
  dds_sub <- dds[, keep]

  # Drop unused factor levels  ✅ FIX FOR droplevels ERROR
  colData(dds_sub)[[group_col]] <- droplevels(factor(colData(dds_sub)[[group_col]]))
  colData(dds_sub)[[age_col]]   <- droplevels(factor(colData(dds_sub)[[age_col]]))

  # Set control as reference
  colData(dds_sub)[[group_col]] <- relevel(colData(dds_sub)[[group_col]], ref = control_group)

  # -------------------------
  # Design
  # -------------------------
  design(dds_sub) <- as.formula(paste("~", age_col, "+", group_col, "+", paste0(age_col, ":", group_col)))

  dds_sub <- DESeq(dds_sub)
  print(resultsNames(dds_sub))

  rn <- resultsNames(dds_sub)

  age_name <- rn[grepl(paste0("^", age_col, "_"), rn)]
  interaction_name <- rn[grepl(paste0(age_col, ".*", group_col), rn)]


  # -------------------------
  # Results
  # -------------------------
  res_ctrl_age <- results(dds_sub, name = age_name)

  res_interaction <- results(dds_sub, name = interaction_name)

  res_test_age <- results(dds_sub, contrast = list(c(age_name, interaction_name)))

  return(list(
    dds = dds_sub,
    res_ctrl_age = res_ctrl_age,
    res_test_age = res_test_age,
    res_interaction = res_interaction,
    coef_names = rn
  ))
}


```

Helper Function that gets all significant genes with padj < 0.05
```{r}
get_sig_genes <- function(res, alpha = 0.05) {
  res <- res[order(res$padj), ]
  sig <- subset(res, !is.na(padj) & padj < alpha)

  list(
    all  = sig,
    up   = sig[sig$log2FoldChange > 0, ],
    down = sig[sig$log2FoldChange < 0, ]
  )
}

```


Now using this function DESeq function
```{r}
table(dds_filt$Group, dds_filt$Age)

out_WSPDsa <- run_age_comparison(dds,control_group = "FAPDsa", test_group = "WSSCsa")
```

```{r}
sig_test_age <- get_sig_genes(out_WSPDsa$res_test_age)
sig_interact <- get_sig_genes(out_WSPDsa$res_interaction)

```



Exporting Top Control Genes to Excel
```{r}
wb <- createWorkbook()

sig_out  <- cbind(Gene = rownames(sig), sig)
up_out   <- cbind(Gene = rownames(up), up)
down_out <- cbind(Gene = rownames(down), down)

addWorksheet(wb, "All_significant")
addWorksheet(wb, "Upregulated_28d")
addWorksheet(wb, "Downregulated_28d")

writeData(wb, "All_significant", sig_out, rowNames = TRUE)
writeData(wb, "Upregulated_28d", up_out, rowNames = TRUE)
writeData(wb, "Downregulated_28d", down_out, rowNames = TRUE)

saveWorkbook(wb, "SignificantAgeControlGenes.xlsx", overwrite = TRUE)

```



Heat map of the most significant genes
```{r}
vsd_ctrl <- vst(dds_ctrl, blind = FALSE)
plotPCA(vsd_ctrl, intgroup = "Age")

topgenes <- head(rownames(res_ctrl_time), 15)
mat_ctrl <- assay(vsd_ctrl)[topgenes, ]
pheatmap::pheatmap(mat_ctrl, scale = "row",
                   annotation_col = as.data.frame(colData(vsd_ctrl)[, "Age", drop=FALSE]))

```

Gigantic Helper Function Test that finds all significant genes from the DESeq Object created, and then runs KEGG and GO enrichment, putting all into a dataframe before returning that dataframe.
```{r}
run_enrichment_pipeline <- function(deseq_list, 
                                    alpha = 0.05,
                                    orgdb = org.Mm.eg.db,
                                    organism = "mmu",
                                    fromType = "ENSEMBL",
                                    keyType = "ENTREZID",
                                    go_ont = "BP",
                                    pval_cutoff = 0.05,
                                    qval_cutoff = 0.05,
                                    strip_version = TRUE) {
  
  # ----------------------------
  # Helper to extract significant genes
  # ----------------------------
  get_sig_genes <- function(res, alpha = 0.05) {
    res <- res[order(res$padj), ]
    sig <- subset(res, !is.na(padj) & padj < alpha)
    list(
      all  = sig,
      up   = sig[sig$log2FoldChange > 0, ],
      down = sig[sig$log2FoldChange < 0, ]
    )
  }
  
  # ----------------------------
  # Helper to run GO/KEGG enrichment
  # ----------------------------
  run_GO_KEGG <- function(up_genes, down_genes) {
    # strip version numbers if needed
    genes_up   <- if(strip_version) sub("\\..*", "", rownames(up_genes)) else rownames(up_genes)
    genes_down <- if(strip_version) sub("\\..*", "", rownames(down_genes)) else rownames(down_genes)
    
    # Convert IDs
    up_ids <- bitr(genes_up, fromType = fromType, toType = keyType, OrgDb = orgdb)
    down_ids <- bitr(genes_down, fromType = fromType, toType = keyType, OrgDb = orgdb)
    
    # GO
    ego_up <- enrichGO(gene = up_ids$ENTREZID,
                       OrgDb = orgdb,
                       keyType = keyType,
                       ont = go_ont,
                       pAdjustMethod = "BH",
                       pvalueCutoff = pval_cutoff,
                       qvalueCutoff = qval_cutoff)
    
    ego_down <- enrichGO(gene = down_ids$ENTREZID,
                         OrgDb = orgdb,
                         keyType = keyType,
                         ont = go_ont,
                         pAdjustMethod = "BH",
                         pvalueCutoff = pval_cutoff,
                         qvalueCutoff = qval_cutoff)
    
    # KEGG
    ekegg_up <- enrichKEGG(gene = up_ids$ENTREZID,
                           organism = organism,
                           pvalueCutoff = pval_cutoff)
    
    ekegg_down <- enrichKEGG(gene = down_ids$ENTREZID,
                             organism = organism,
                             pvalueCutoff = pval_cutoff)
    
    list(GO_up = ego_up, GO_down = ego_down, KEGG_up = ekegg_up, KEGG_down = ekegg_down)
  }
  
  # ----------------------------
  # 1. Extract sig genes for test_age and interaction
  # ----------------------------
  sig_test_age <- get_sig_genes(deseq_list$res_test_age, alpha)
  sig_interaction <- get_sig_genes(deseq_list$res_interaction, alpha)
  
  # ----------------------------
  # 2. Run enrichment
  # ----------------------------
  enrich_test_age <- run_GO_KEGG(sig_test_age$up, sig_test_age$down)
  enrich_interaction <- run_GO_KEGG(sig_interaction$up, sig_interaction$down)
  
  # ----------------------------
  # 3. Build summary dataframe of all significant genes
  # ----------------------------
  # Helper to make df
  make_df <- function(sig_list, label) {
    df_up <- data.frame(
      Gene = rownames(sig_list$up),
      log2FC = sig_list$up$log2FoldChange,
      padj = sig_list$up$padj,
      direction = "up",
      contrast = label
    )
    df_down <- data.frame(
      Gene = rownames(sig_list$down),
      log2FC = sig_list$down$log2FoldChange,
      padj = sig_list$down$padj,
      direction = "down",
      contrast = label
    )
    rbind(df_up, df_down)
  }
  
  df_test_age <- make_df(sig_test_age, "TestAge")
  df_interaction <- make_df(sig_interaction, "Interaction")
  
  all_sig_genes_df <- rbind(df_test_age, df_interaction)
  
  # ----------------------------
  # 4. Return everything
  # ----------------------------
  list(
    sig_test_age = sig_test_age,
    sig_interaction = sig_interaction,
    enrich_test_age = enrich_test_age,
    enrich_interaction = enrich_interaction,
    all_sig_genes_df = all_sig_genes_df
  )
}

```


Testing the significant gene along with KEGG and GO enrichment function.
```{r}
pipeline_results <- run_enrichment_pipeline(out_WSPDsa)


```

  

New Helper Function for getting all the plots of KEGG and GO enrichment
```{r}
plot_enrichment_dot_barplots <- function(pipeline_results,
                                         control_group = "FAPDsa",
                                         test_group    = "WSSCsa",
                                         showCategoryGO = 10,
                                         showCategoryKEGG = 10) {
  
  plots <- list()
  
  # ----------------------------
  # Helper function to generate both dotplot and barplot
  # ----------------------------
  make_plots <- function(enrich_obj, title_prefix, showCategory) {
    out <- list()
    out[[paste0(title_prefix, "_dot")]] <- dotplot(enrich_obj, showCategory = showCategory) +
      ggtitle(paste0(title_prefix, " - Dotplot"))
    out[[paste0(title_prefix, "_bar")]] <- barplot(enrich_obj, showCategory = showCategory) +
      ggtitle(paste0(title_prefix, " - Barplot"))
    return(out)
  }
  
  # ----------------------------
  # 1. Test group aging
  # ----------------------------
  if(!is.null(pipeline_results$enrich_test_age$GO_up)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_test_age$GO_up, 
                                 paste0(test_group, " Aging - GO Up"), showCategoryGO))
  }
  if(!is.null(pipeline_results$enrich_test_age$GO_down)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_test_age$GO_down, 
                                 paste0(test_group, " Aging - GO Down"), showCategoryGO))
  }
  if(!is.null(pipeline_results$enrich_test_age$KEGG_up)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_test_age$KEGG_up, 
                                 paste0(test_group, " Aging - KEGG Up"), showCategoryKEGG))
  }
  if(!is.null(pipeline_results$enrich_test_age$KEGG_down)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_test_age$KEGG_down, 
                                 paste0(test_group, " Aging - KEGG Down"), showCategoryKEGG))
  }
  
  # ----------------------------
  # 2. Interaction (aging difference vs control)
  # ----------------------------
  if(!is.null(pipeline_results$enrich_interaction$GO_up)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_interaction$GO_up, 
                                 paste0(test_group, " vs ", control_group, " - GO Up"), showCategoryGO))
  }
  if(!is.null(pipeline_results$enrich_interaction$GO_down)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_interaction$GO_down, 
                                 paste0(test_group, " vs ", control_group, " - GO Down"), showCategoryGO))
  }
  if(!is.null(pipeline_results$enrich_interaction$KEGG_up)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_interaction$KEGG_up, 
                                 paste0(test_group, " vs ", control_group, " - KEGG Up"), showCategoryKEGG))
  }
  if(!is.null(pipeline_results$enrich_interaction$KEGG_down)) {
    plots <- c(plots, make_plots(pipeline_results$enrich_interaction$KEGG_down, 
                                 paste0(test_group, " vs ", control_group, " - KEGG Down"), showCategoryKEGG))
  }
  
  return(plots)
}

```

Run Controls Graphs so we can compare
```{r}

# Example: control_group = "FAPDsa"
dds_ctrl <- dds_filt[, dds_filt$Group == "FAPDsa"]

# Update design for aging only
design(dds_ctrl) <- ~ Age

# Run DESeq
dds_ctrl <- DESeq(dds_ctrl)

# ===============================
# 2. Extract differential expression for aging in control
# ===============================
res_ctrl_age <- results(dds_ctrl, contrast = c("Age", "28 day", "1 day"))
res_ctrl_age <- res_ctrl_age[order(res_ctrl_age$padj), ]

# Subset significant genes
sig_ctrl <- subset(res_ctrl_age, padj < 0.05)
up_ctrl   <- sig_ctrl[sig_ctrl$log2FoldChange > 0, ]
down_ctrl <- sig_ctrl[sig_ctrl$log2FoldChange < 0, ]

# ===============================
# 3. Run GO/KEGG enrichment for control aging
# ===============================
# Reusing the helper function
ctrl_enrich <- run_GO_KEGG(up_ctrl, down_ctrl)

# ===============================
# 4. Plot dotplots for GO/KEGG
# ===============================
ctrl_plots <- list()

if(!is.null(ctrl_enrich$GO_up)) {
  ctrl_plots$GO_Up_Dot <- dotplot(ctrl_enrich$GO_up, showCategory = 10) +
    ggtitle("Control Aging - GO Up")
}
if(!is.null(ctrl_enrich$GO_down)) {
  ctrl_plots$GO_Down_Dot <- dotplot(ctrl_enrich$GO_down, showCategory = 10) +
    ggtitle("Control Aging - GO Down")
}
if(!is.null(ctrl_enrich$KEGG_up)) {
  ctrl_plots$KEGG_Up_Dot <- dotplot(ctrl_enrich$KEGG_up, showCategory = 10) +
    ggtitle("Control Aging - KEGG Up")
}
if(!is.null(ctrl_enrich$KEGG_down)) {
  ctrl_plots$KEGG_Down_Dot <- dotplot(ctrl_enrich$KEGG_down, showCategory = 10) +
    ggtitle("Control Aging - KEGG Down")
}

if(!is.null(ctrl_enrich$GO_up)) {
  ctrl_plots$GO_Up_Bar <- barplot(ctrl_enrich$GO_up, showCategory = 10) +
    ggtitle("Control Aging - GO Up")
}
if(!is.null(ctrl_enrich$GO_down)) {
  ctrl_plots$GO_Down_Bar <- barplot(ctrl_enrich$GO_down, showCategory = 10) +
    ggtitle("Control Aging - GO Down")
}
if(!is.null(ctrl_enrich$KEGG_up)) {
  ctrl_plots$KEGG_Up_Bar <- barplot(ctrl_enrich$KEGG_up, showCategory = 10) +
    ggtitle("Control Aging - KEGG Up")
}
if(!is.null(ctrl_enrich$KEGG_down)) {
  ctrl_plots$KEGG_Down_Bar <- barplot(ctrl_enrich$KEGG_down, showCategory = 10) +
    ggtitle("Control Aging - KEGG Down")
}

```
Control Plots
```{r}

ctrl_plots$GO_Up_Dot  # example
ctrl_plots$GO_Down_Dot
ctrl_plots$KEGG_Up_Bar
ctrl_plots$KEGG_Down_Bar
```



Now workflow for Gene Comparisons
```{r}
#Defining what Groups we are comparing
out <- run_age_comparison(dds, control_group = "FAPDsa", test_group = "WSSCsa")

#Extract Sig Genes and Run KEGG and GO Enrichment
pipeline_results <- run_enrichment_pipeline_with_universe(out)

# Plotting
plots <- plot_enrichment_dotplots(
  pipeline_results,
  control_group = "FAPDsa",
  test_group    = "WSSCsa",
  showCategoryGO = 10,
  showCategoryKEGG = 10
)

```

```{r}
plots$TestAge_KEGG_Up
plots$Interaction_GO_Up
plots$Interaction_KEGG_Up
plots$Interaction_KEGG_Down
```

Alright so we want to compare FA vs WS, SA vs RS, and then finally culuminate the 2 into FA SA vs WS RS
```{r}
out <- run_age_comparison(dds, control_group = "FAPDsa", test_group = "FAPDrs")

#Extract Sig Genes and Run KEGG and GO Enrichment
pipeline_results <- run_enrichment_pipeline_with_universe(out)

# Plotting
plots <- plot_enrichment_dotplots(
  pipeline_results,
  control_group = "FAPDsa",
  test_group    = "FAPDrs",
  showCategoryGO = 10,
  showCategoryKEGG = 10
)

```
Plots for FA Resolvin and Differences in Aging
```{r}
plots$TestAge_GO_Up
plots$TestAge_GO_Down
plots$TestAge_KEGG_Up
plots$TestAge_KEGG_Down
```

Plots for FA Saline vs FA Resolvin and then signficant genes
```{r}
plots$Interaction_GO_Up
plots$Interaction_GO_Down
plots$Interaction_KEGG_Up
plots$Interaction_KEGG_Down
```
FA Saline vs WS resolvin
```{r}
out <- run_age_comparison(dds_filt, control_group = "FAPDsa", test_group = "WSPDrs")

#Extract Sig Genes and Run KEGG and GO Enrichment
pipeline_results <- run_enrichment_pipeline_with_universe(out)

# Plotting
plots <- plot_enrichment_dot_barplots(
  pipeline_results,
  control_group = "FAPDsa",
  test_group    = "WSPDrs",
  showCategoryGO = 10,
  showCategoryKEGG = 10
)

```
WS Resolvin Plots
```{r}
plots$TestAge_GO_Up
plots$TestAge_GO_Down
plots$TestAge_KEGG_Up
plots$TestAge_KEGG_Down
```

FA Saline vs WS Resolvin Plots
```{r}
plots$Interaction_GO_Up
plots$Interaction_GO_Down
plots$Interaction_KEGG_Up
plots$Interaction_KEGG_Down
```

Testing Enrichment pipeline with Universe of Genes
```{r}
run_enrichment_pipeline_with_universe <- function(deseq_list, 
                                                  alpha = 0.05,
                                                  orgdb = org.Mm.eg.db,
                                                  organism = "mmu",
                                                  fromType = "ENSEMBL",
                                                  keyType = "ENTREZID",
                                                  go_ont = "BP",
                                                  pval_cutoff = 0.05,
                                                  qval_cutoff = 0.05,
                                                  strip_version = TRUE,
                                                  define_universe = TRUE) {
  
  # ----------------------------
  # Helper to extract significant genes
  # ----------------------------
  get_sig_genes <- function(res, alpha = 0.05) {
    res <- res[order(res$padj), ]
    sig <- subset(res, !is.na(padj) & padj < alpha)
    list(
      all  = sig,
      up   = sig[sig$log2FoldChange > 0, ],
      down = sig[sig$log2FoldChange < 0, ]
    )
  }
  
  # ----------------------------
  # Helper to run GO/KEGG enrichment
  # ----------------------------
  run_GO_KEGG <- function(up_genes, down_genes, universe_ids = NULL) {
    genes_up   <- if(strip_version) sub("\\..*", "", rownames(up_genes)) else rownames(up_genes)
    genes_down <- if(strip_version) sub("\\..*", "", rownames(down_genes)) else rownames(down_genes)
    
    up_ids <- bitr(genes_up, fromType = fromType, toType = keyType, OrgDb = orgdb)
    down_ids <- bitr(genes_down, fromType = fromType, toType = keyType, OrgDb = orgdb)
    
    ego_up <- enrichGO(gene = up_ids$ENTREZID,
                       OrgDb = orgdb,
                       keyType = keyType,
                       ont = go_ont,
                       universe = universe_ids,
                       pAdjustMethod = "BH",
                       pvalueCutoff = pval_cutoff,
                       qvalueCutoff = qval_cutoff)
    
    ego_down <- enrichGO(gene = down_ids$ENTREZID,
                         OrgDb = orgdb,
                         keyType = keyType,
                         ont = go_ont,
                         universe = universe_ids,
                         pAdjustMethod = "BH",
                         pvalueCutoff = pval_cutoff,
                         qvalueCutoff = qval_cutoff)
    
    ekegg_up <- enrichKEGG(gene = up_ids$ENTREZID,
                           organism = organism,
                           universe = universe_ids,
                           pvalueCutoff = pval_cutoff)
    
    ekegg_down <- enrichKEGG(gene = down_ids$ENTREZID,
                             organism = organism,
                             universe = universe_ids,
                             pvalueCutoff = pval_cutoff)
    
    list(GO_up = ego_up, GO_down = ego_down, KEGG_up = ekegg_up, KEGG_down = ekegg_down)
  }
  
  # ----------------------------
  # 0. Define universe if requested
  # ----------------------------
  if (define_universe) {
    all_genes <- rownames(deseq_list$res_test_age)
    if (strip_version) {
      all_genes <- sub("\\..*", "", all_genes)
    }
    universe_ids <- bitr(all_genes, fromType = fromType, toType = keyType, OrgDb = orgdb)$ENTREZID
  } else {
    universe_ids <- NULL
  }
  
  # ----------------------------
  # 1. Extract sig genes
  # ----------------------------
  sig_test_age <- get_sig_genes(deseq_list$res_test_age, alpha)
  sig_interaction <- get_sig_genes(deseq_list$res_interaction, alpha)
  
  # ----------------------------
  # 2. Run enrichment using universe
  # ----------------------------
  enrich_test_age <- run_GO_KEGG(sig_test_age$up, sig_test_age$down, universe_ids)
  enrich_interaction <- run_GO_KEGG(sig_interaction$up, sig_interaction$down, universe_ids)
  
  # ----------------------------
  # 3. Build summary dataframe
  # ----------------------------
  make_df <- function(sig_list, label) {
    df_up <- data.frame(
      Gene = rownames(sig_list$up),
      log2FC = sig_list$up$log2FoldChange,
      padj = sig_list$up$padj,
      direction = "up",
      contrast = label
    )
    df_down <- data.frame(
      Gene = rownames(sig_list$down),
      log2FC = sig_list$down$log2FoldChange,
      padj = sig_list$down$padj,
      direction = "down",
      contrast = label
    )
    rbind(df_up, df_down)
  }
  
  df_test_age <- make_df(sig_test_age, "TestAge")
  df_interaction <- make_df(sig_interaction, "Interaction")
  all_sig_genes_df <- rbind(df_test_age, df_interaction)
  
  # ----------------------------
  # 4. Return everything
  # ----------------------------
  list(
    sig_test_age = sig_test_age,
    sig_interaction = sig_interaction,
    enrich_test_age = enrich_test_age,
    enrich_interaction = enrich_interaction,
    all_sig_genes_df = all_sig_genes_df,
    gene_universe = universe_ids
  )
}

```

Redoing PCAs and keeping Consistent Layouts
```{r}

pca_input <- t(vst_filt)

any(apply(pca_input, 2, var) == 0)


pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  metadata[rownames(pca$x), ]
)

pve <- pca$sdev^2 / sum(pca$sdev^2)
pve[1:5]


plot(
  pca_df$PC1, pca_df$PC2,
  xlab = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
  ylab = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
  pch = 19
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)

print(pca_df$Outlier)

pca_df$GroupFactor <- meta[rownames(pca_df), "Group"]
#pca_df$GroupFactor <- factor(pca_df$GroupID)


ggplot(pca_df, aes(x = PC1, y = PC2, color = GroupFactor)) +
   geom_point(aes(fill = GroupFactor),
             shape = 21,
             color = "black",
             stroke = 0.8,
             size = 6) +
    geom_text(
    data = subset(pca_df, Outlier),
    aes(label =  sub(".*-", "", Sample)),
    size = 4,
    color="black"
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of 8 Groups",
    subtitle="DESeq Filtered Beforehand"
  ) +
  theme_classic()


```



Doing some experiments and looking at top genes considering every factor now
```{r}
coefs <- resultsNames(dds)
coefs <- coefs[coefs != "Intercept"]

head(coefs)

res_list <- lapply(coefs, function(cn) {
  res <- results(dds, name = cn)
  res <- as.data.frame(res)
  res$ENSEMBL <- rownames(res)
  
  sig <- res[!is.na(res$padj) & res$padj < 0.05, ]
  return(sig)
})

names(res_list) <- coefs

all_sig <- unique(unlist(lapply(res_list, rownames)))


```

```{r}
res <- results(dds, name="Exposure_WS_vs_FA")
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)
head(res)

sig <- res[!is.na(res$padj) & res$padj <0.05, ]

# Add direction column: Up / Down
sig$direction <- ifelse(sig$log2FoldChange > 0, "Up", "Down")
table(sig$direction)

sig$ENTREZID <- ids$ENTREZID[match(str_remove(sig$ENSEMBL, "\\.\\d+$"), ids$clean_ensembl)]
sig_entrez <- unique(sig$ENTREZID)
sig_entrez <- sig_entrez[!is.na(sig_entrez)]

sig_up_entrez   <- sig$ENTREZID[sig$direction == "Up"]
sig_up_entrez   <- unique(sig_up_entrez[!is.na(sig_up_entrez)])

sig_down_entrez <- sig$ENTREZID[sig$direction == "Down"]
sig_down_entrez <- unique(sig_down_entrez[!is.na(sig_down_entrez)])

```

```{r}

  # remove NAs

# Universe / background: all tested genes with Entrez IDs
bg_entrez <- unique(ids$ENTREZID)
bg_entrez <- bg_entrez[!is.na(bg_entrez)]

ego_up <- enrichGO(
  gene          = sig_up_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

ego_down <- enrichGO(
  gene          = sig_down_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)


ego_up_df   <- as.data.frame(ego_up) %>% mutate(direction = "Up")
ego_down_df <- as.data.frame(ego_down) %>% mutate(direction = "Down")

ego_combined <- bind_rows(ego_up_df, ego_down_df)

```

```{r}
dotplot(ego_up, showCategory=10,title="GO Up")
dotplot(ego_down, showCategory=10, title="GO Down")
```

Want to do GO enrichment and find top 20 genes of true control 1d vs WSPDsa 1d
```{r}
meta_test <- meta[
  meta$Age == "1 day" &
  meta$Treatment == "sa" &
  meta$Diet == "PD" &
  meta$Exposure %in% c("FA", "WS"),
]

table(meta_test$Exposure, meta_test$Age)
meta_test

count_test <- count_matrix[,rownames(meta_test)]

dds <- DESeqDataSetFromMatrix(
  countData = count_test,
  colData   = meta_test,
  design    = ~ Group
)

#Potential Filtering Option
dds <- dds[rowSums(counts(dds)) >= 10, ]
dds <- DESeq(dds)

resultsNames(dds)

res <- results(dds)
head(res)
```

```{r}
res <- results(dds, name = "Group_WSPDsa_vs_FAPDsa")
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

# keep only significant genes
sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

# add absolute log2 fold change
sig$absLog2FC <- abs(sig$log2FoldChange)

# sort by absolute effect size
sig_sorted <- sig[order(-sig$absLog2FC), ]
sig_sorted$clean_ensembl <- str_remove(sig_sorted$ENSEMBL, "\\.\\d+$")
sig_sorted$SYMBOL <- ids$SYMBOL[match(sig_sorted$clean_ensembl, ids$clean_ensembl)]

# Only get mapped symbols
sig_mapped <- sig_sorted[!is.na(sig_sorted$SYMBOL) & sig_sorted$SYMBOL != "", ]

# look at top hits
head(sig_sorted, 10)

```

```{r}
# take top 20 by absolute log2FC
top20 <- head(sig_mapped, 20)

# clean Ensembl again just in case
top20$clean_ensembl <- str_remove(top20$ENSEMBL, "\\.\\d+$")

# map gene symbols
top20$SYMBOL <- ids$SYMBOL[match(top20$clean_ensembl, ids$clean_ensembl)]

# if some symbols are missing, fall back to Ensembl
top20$LABEL <- ifelse(is.na(top20$SYMBOL), top20$clean_ensembl, top20$SYMBOL)

# classify direction
top20$Direction <- ifelse(top20$log2FoldChange > 0, "Upregulated in WSPDsa", "Downregulated in WSPDsa")


ggplot(top20, aes(x = reorder(LABEL, absLog2FC),
                  y = absLog2FC,
                  fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6))+
  labs(
    x = "Gene Name",
    y = "| log2(FoldChange) |",
    title = "Top 20 Log Fold Genes (Mapped Gene Symbols Only)",
    subtitle = "WSPDsa 1 Day vs True Control (FAPDsa 1 Day)"
  ) +
  theme_classic(base_size = 13)

```

```{r}
sig$clean_ensembl <- str_remove(sig$ENSEMBL, "\\.\\d+$")
sig$ENTREZID <- ids$ENTREZID[match(sig$clean_ensembl, ids$clean_ensembl)]

# remove genes without Entrez IDs
sig <- sig[!is.na(sig$ENTREZID), ]

sig_up   <- sig[sig$log2FoldChange > 0, ]
sig_down <- sig[sig$log2FoldChange < 0, ]

up_entrez   <- unique(sig_up$ENTREZID)
down_entrez <- unique(sig_down$ENTREZID)

# background universe
bg_entrez <- unique(ids$ENTREZID)
bg_entrez <- bg_entrez[!is.na(bg_entrez)]

go_up <- enrichGO(
  gene          = up_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

go_down <- enrichGO(
  gene          = down_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)
```

```{r}
go_up_df <- as.data.frame(go_up)

# sort by Count descending
go_up_df <- go_up_df[order(-go_up_df$Count), ]

# pick top N categories
topN <- 15
go_up_top <- go_up_df[1:topN, ]


ggplot(go_up_top, aes(x = reorder(Description, Count), y = Count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "GO Term",
    y = "Gene Count",
    title = "Top 15 GO Terms – Upregulated in WSPDsa 1 Day"
  ) +
  theme_minimal(base_size = 13)

go_down_df <- as.data.frame(go_down)
go_down_df <- go_down_df[order(-go_down_df$Count), ]
go_down_top <- go_down_df[1:15, ]

ggplot(go_down_top, aes(x = reorder(Description, Count), y = Count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "GO Term",
    y = "Gene Count",
    title = "Top 15 GO Terms – Downregulated in WSPDsa 1 Day"
  ) +
  theme_minimal(base_size = 13)


```


Recreating Jennifers Poster Top 5 Genes Graph
Please Run DDS with all the differing conditions
```{r}
resultsNames(dds)
terms <- setdiff(resultsNames(dds),"Intercept")

terms[1]

res <- results(dds, name = terms[1])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

# List of Genes that we will exclude from the top genes list
exclude <- res[!is.na(res$padj) & res$padj < 0.05, ]

# List of Woodsmoke Genes that are significant
res <- results(dds, name = terms[3])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

# Filtering for only top genes in exposure
exposure_genes <- sig[!sig$ENSEMBL %in% exclude$ENSEMBL, ]

res <- results(dds, name = terms[2])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

common_genes <- intersect(exposure_genes$ENSEMBL,sig$ENSEMBL)
exposure_genes <- exposure_genes[exposure_genes$ENSEMBL %in% common_genes, ]
exposure_genes <- exposure_genes[order(exposure_genes$padj),]

# Now we need to map these to their according gene names
exposure_genes$clean_ensembl <- str_remove(exposure_genes$ENSEMBL, "\\.\\d+$")
exposure_genes$SYMBOL <- ids$SYMBOL[match(exposure_genes$clean_ensembl, ids$clean_ensembl)]


exposure_genes <- exposure_genes[!is.na(exposure_genes$SYMBOL) & exposure_genes$SYMBOL != "", ]
head(exposure_genes$SYMBOL)

# Now we just need to plot all of this with our different groups
expr <- assay(vsd)
rownames(expr) <- str_remove(rownames(expr), "\\.\\d+$")

expr <- expr[ids$clean_ensembl[ids$SYMBOL %in% exposure_genes$SYMBOL[1:5]],,drop=FALSE]
rownames(expr) <- ids$SYMBOL[match(rownames(expr),ids$clean_ensembl)]

expr_long <- as.data.frame(expr) %>%
  mutate(SYMBOL = rownames(.)) %>%
  pivot_longer(
    cols = -SYMBOL,
    names_to = "sample",
    values_to = "expression"
  )

expr_long <- expr_long %>%
  left_join(meta, by = c("sample" = "SampleID"))

meta$plot_group

groups_keep <- c(
  "FAPDsa 1 day",
  "FAWDsa 1 day",
  "WSPDsa 1 day",
  "WSWDsa 1 day",
  "FAPDsa 28 day",
  "WSPDsa 28 day",
  "FAWDsa 28 day",
  "WSWDsa 28 day"
)

expr_long <- expr_long %>%
  filter(plot_group %in% groups_keep)

expr_long$plot_group <- factor(
  expr_long$plot_group,
  levels = groups_keep
)

options(
  repr.plot.width  = 14,
  repr.plot.height = 8
)

ggplot(expr_long, aes(x = plot_group, y = expression, color = plot_group)) +
  stat_summary(
  fun = mean,
  geom = "crossbar",
  width = 0.6,
  color = "black",
  linewidth = 0.5
) +
geom_jitter(
  width = 0.15,
  size = 3,
  alpha = 0.8
) +
  facet_wrap(~ SYMBOL, scales = "free_y") +
  theme_classic(base_size = 13) +
   theme(
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.5),
    panel.grid.minor.y = element_line(color = "grey90", linewidth = 0.25),
    panel.grid.major.x = element_line(color = "grey88", linewidth = 0.4),
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1.09, hjust = 1),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = c(1, 0),      # bottom right
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "black"),  # optional border
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11)
  )  +
  labs(
    x = "Exposure / Age Group",
    y = "VST Normalized Expression",
    color = "Group",
    title = "Top 5 Exposure-Specific Genes"
  )

head(exposure_genes)
```

```{r}

top_genes_all <- lapply(
  setdiff(resultsNames(dds), "Intercept"),
  function(coef_name) {

    res <- results(dds, name = coef_name) %>%
      as.data.frame() %>%
      rownames_to_column("ENSEMBL") %>%
      filter(!is.na(padj), padj < 0.05) %>%
      arrange(desc(abs(log2FoldChange))) %>%
      slice_head(n = 10) %>%
      mutate(term = coef_name)

    res
  }
)

top_genes_all <- bind_rows(top_genes_all)

print(top_genes_all)
top_genes_all$clean_ensembl <- str_remove(top_genes_all$ENSEMBL, "\\.\\d+$")
top_genes_all$SYMBOL <- ids$SYMBOL[match(top_genes_all$clean_ensembl, ids$clean_ensembl)]

top_genes_all <- top_genes_all[!is.na(top_genes_all$SYMBOL) & top_genes_all$SYMBOL != "", ]
print(top_genes_all$SYMBOL)

vst_sub <- vst_df[rownames(vst_df) %in% top_genes_all$ENSEMBL, , drop = FALSE]

rownames(vst_sub) <- top_genes_all$SYMBOL[
  match(rownames(vst_sub), top_genes_all$ENSEMBL)
]

vst_scaled <- t(scale(t(vst_sub)))

vst_long <- vst_sub %>%
  as.data.frame() %>%
  mutate(SYMBOL = rownames(.)) %>%
  pivot_longer(
    cols = -SYMBOL,
    names_to = "SampleID",
    values_to = "expression"
  )

head(meta)
meta$plot_group
# Join metadata (plot_group and groupID)

vst_long <- vst_long %>%
  mutate(SampleID_numeric = as.character(as.integer(str_remove(SampleID, "^Yim-"))))


vst_long <- vst_long %>%
  left_join(dplyr::select(meta, SampleID, plot_group, GroupID), 
            by = c("SampleID_numeric" = "SampleID"))


print(head(vst_long))

avg_df <- vst_long %>%
  group_by(SYMBOL, plot_group) %>%
  summarise(avg_expression = mean(expression), .groups = "drop")

avg_mat <- avg_df %>%
  pivot_wider(names_from = plot_group, values_from = avg_expression) %>%
  column_to_rownames("SYMBOL") %>%
  as.matrix()

avg_scaled <- t(scale(t(avg_mat)))


pheatmap(avg_scaled)


```
Trying out umap for maybe different visualisation techniques
```{r}
vst_umap <- umap(vst)

```

```{r}
#head(vst_umap)

umap_df <- as.data.frame(vst_umap$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

#umap_df$SampleID <- colnames(vst)
head(umap_df)

umap_df$ensembl <- rownames(umap_df)


print(summary(umap_df))

plot(umap_df$UMAP1,umap_df$UMAP2)

# Need to look into classification of genes into potential categories 
# Otherwise cannot look into this one
```

Sample Level UMAp
```{r}
vst_t <- t(vst)

vst_umap_samples <- umap(vst_t, n_components = 3, random_state=52)

```

Analysis
```{r}

head(rownames(vst_umap_samples$layout))
# Example output: "Yim-001", "Yim-002", ...

# Make meta$SampleID match this format
meta$SampleID <- paste0("Yim-", sprintf("%03d", 1:nrow(meta)))

# Now create umap_sample_df
umap_sample_df <- as.data.frame(vst_umap_samples$layout)
colnames(umap_sample_df) <- c("UMAP1", "UMAP2","UMAP3")
umap_sample_df$SampleID <- rownames(umap_sample_df)

# Merge by SampleID
umap_sample_df <- merge(umap_sample_df, meta, by = "SampleID")

head(umap_sample_df)

ggplot(umap_sample_df, aes(x = UMAP1, y = UMAP2, color = plot_group0055548487878787)) +
  geom_point(size = 3) +
  theme_minimal()
```

```{r}
plot_ly(umap_sample_df, x=~UMAP1,y=~UMAP2,z=~UMAP3,color=~Group,type="scatter3d",mode="markers",marker=list(size=4))
```


Looking at GSEA now
```{r}
res <- results(dds, name="Exposure_WS_vs_FA")
res <- res[!is.na(res$stat),]

gene_ranks <- res$stat
names(gene_ranks) <- rownames(res)

gene_ranks <- sort(gene_ranks, decreasing = TRUE)


gene_df <- data.frame(
  ensembl_ids = names(gene_ranks),
  stat = gene_ranks,
  stringsAsFactors = FALSE
)

gene_df$clean_ensembl <- str_remove(gene_df$ensembl_ids,"\\.\\d+$")

gene_df$ENTREZID <- ids$ENTREZID[match(gene_df$clean_ensembl, ids$clean_ensembl)]

gene_df <- gene_df[!is.na(gene_df$ENTREZID) & gene_df$ENTREZID != "",]

# Order by absolute stat (largest effect first)
gene_df <- gene_df[order(abs(gene_df$stat), decreasing = TRUE), ]

# Remove duplicate ENTREZ IDs (keep first = strongest signal)
gene_df <- gene_df[!duplicated(gene_df$ENTREZID), ]

gene_ranks <- gene_df$stat
names(gene_ranks) <- gene_df$ENTREZID

gene_ranks <- sort(gene_ranks, decreasing = TRUE)


```

After cleaning and prepping our thing, now we can do gsea
```{r}
gsea_res <- gseGO(
  geneList = gene_ranks,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "ENTREZID"
)

```
Visualizations now
```{r}
dotplot(gsea_res, showCategory = 10)

gsea_df <- as.data.frame(gsea_res)
head(gsea_df)

# GSeaplot analysis for specific geneSETIDs
gseaplot2(gsea_res, geneSetID = "GO:0006631")

```





















