---
title: "RNA Seq"
output: html_notebook
---

# First Load All Libraries
```{r}
library(limma)
library(DESeq2)
library(ggplot2)
library(tidyverse)
library(readr)
library(ggrepel)
library(clusterProfiler)
library(org.Mm.eg.db)   # change to org.Hs.eg.db if human
library(enrichplot)
library(DOSE)
library(readxl)
library(AnnotationDbi)
library(stringr)
library(dplyr)
library(pheatmap)
library(umap)
library(plotly)
library(fgsea)
library(caret)
library(glmnet)
library(xgboost)
library(WGCNA)
allowWGCNAThreads()
library(igraph)
library(openxlsx)
```

```{r}
theme_set(
  theme_classic(base_size = 14, base_family = "Arial") +
    theme(
      plot.title = element_text(face = "bold"),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12),
      legend.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)
```



```{r}
FastP_featurecounts <- read.csv("data/FastP_featurecounts.csv")
metadata <- read.csv("data/metadata.csv")
```

# Cleaned up Data Matrix
```{r}
head(FastP_featurecounts)

fc <- FastP_featurecounts[,-(2:6)]
colnames(fc) <- as.character(unlist(fc[1,]))
fc <- fc[-1,]
rownames(fc) <- as.character(unlist(fc[,1]))
fc <- fc[,-1]
head(fc)

count_matrix <- as.matrix(fc)
head(count_matrix)
storage.mode(count_matrix) <- "numeric"
print(is.numeric(count_matrix))
```

Adding Column for Ensembl IDs to Gene IDs
```{r}
ids <- data.frame(ensembl_ids = rownames(fc), stringsAsFactors = FALSE)

# 2. Clean Ensembl IDs (remove .version)
ids$clean_ensembl <- str_remove(ids$ensembl_ids, "\\.\\d+$")

# 3. Query annotations
annotations <- AnnotationDbi::select(org.Mm.eg.db,
                      keys = unique(ids$clean_ensembl),
                      columns = c("SYMBOL", "GENENAME", "ENTREZID"),
                      keytype = "ENSEMBL")

# 4. Enforce one-to-one mapping
annot_unique <- annotations[!duplicated(annotations$ENSEMBL), ]
rownames(annot_unique) <- annot_unique$ENSEMBL

# 5. Build master table by matching CLEAN IDs
ids$SYMBOL   <- annot_unique[ids$clean_ensembl, "SYMBOL"]
ids$GENENAME <- annot_unique[ids$clean_ensembl, "GENENAME"]
ids$ENTREZID <- annot_unique[ids$clean_ensembl, "ENTREZID"]

# 6. Inspect
head(ids)
```


# Check Files are Loaded in and begin cleaning of them
```{r}
# Check head of read in files, and then ccreate copies ot modify
head(metadata)
meta <- metadata

# Correct unique grouping so that we can quickly split on it
meta <- meta %>%
  mutate(
    GroupID = group_indices(., Age, Exposure, Diet, Treatment)
  )

meta %>%
  distinct(Age, Exposure, Diet, Treatment, GroupID) %>%
  arrange(GroupID)

meta$SampleID <- rownames(meta)


meta <- meta %>%
  mutate(
    plot_group = paste0(Exposure,Diet,Treatment, " ", Age)
  )

meta <- as.data.frame(meta)
meta$SampleID <- sprintf("Yim-%03d", as.integer(rownames(meta)))
rownames(meta) <- meta$SampleID
meta$SampleID <- NULL

meta

```
Now we want to compute basic DESeq2 Object and Limma Object, with transforms of scale and quantile.
```{r}
# Run DESeq2 Object along with most result important genes, along with variance transform
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData   = metadata,
  design    = ~ Age + Exposure + Diet + Treatment
)

#Potential Filtering Option
dds <- dds[rowSums(counts(dds)) >= 10, ]

dds <- DESeq(dds)


# Getting results of each basic design pattern along
# with VST transform for visualization purposes
# and norm_counts for more indepth looks along with statistical tests
res <- results(dds)
res <- res[order(res$padj),]
head(res)

vsd <- vst(dds, blind=FALSE)
vst <- assay(vsd)

vst_df <- as.data.frame(vst)
head(vst_df)

terms <- resultsNames(dds)
terms <- terms[-1]
print(terms)

norm_counts <- counts(dds,normalized=TRUE)
```
Volcano plots of the deseq2 object
```{r}
resultsNames(dds)
terms <- resultsNames(dds)

res <- results(dds,name=terms[2])

res <- as.data.frame(res)
# Need to clean the res dataframe
res <- res[complete.cases(res[,c("pvalue")]),]
res$color <- "gray"

res$color[res$pvalue < 0.05 & res$log2FoldChange > 1] <- "red"
res$color[res$pvalue < 0.05 & res$log2FoldChange < -1] <- "blue"

sig <- c(sum(res$color == "blue"))
sig[2] <- sum(res$color == "red")
sig[3] <- sum(res$color=="gray")

annot <- data.frame(
  x = c(-10, -10, -10),   # x-position of labels
  y = c(15, 14, 13),  # y-position (stack vertically)
  label = c(
    paste("Down:", sig[1]),
    paste("Up:", sig[2]),
    paste("NS:", sig[3])
  ),
  color = c("blue", "red", "black")
)

head(res,20)

ggplot(res, aes(x=log2FoldChange,y=-log10(pvalue), color=color)) +
  geom_point(size=2.5,alpha=0.85)+
  geom_hline(yintercept=-log10(0.05), linetype="dashed")+
  geom_vline(xintercept=c(-1,1),linetype="dashed")+
  scale_color_identity()+ 
   geom_text(
    data = annot,
    aes(x = x, y = y, label = label, color = color),
    inherit.aes = FALSE,
    size = 5
  )+
  labs(x = "log2FoldChange",
       title=paste0("Volcano Plot of ", terms[2])) +
  theme_classic()

```

Function that given the dds object and experiment name, it will create the volcano plot for any upregulated and downregulated genes in the differential expression.
```{r}
volcano_plot <- function(dds, exp_name){

res <- results(dds,name=exp_name)

res <- as.data.frame(res)
# Need to clean the res dataframe
res <- res[complete.cases(res[,c("pvalue")]),]
res$color <- "gray"

res$color[res$pvalue < 0.05 & res$log2FoldChange > 1] <- "red"
res$color[res$pvalue < 0.05 & res$log2FoldChange < -1] <- "blue"

sig <- c(sum(res$color == "blue"))
sig[2] <- sum(res$color == "red")
sig[3] <- sum(res$color=="gray")

annot <- data.frame(
  x = c(-10, -10, -10),   # x-position of labels
  y = c(15, 14, 13),  # y-position (stack vertically)
  label = c(
    paste("Down:", sig[1]),
    paste("Up:", sig[2]),
    paste("NS:", sig[3])
  ),
  color = c("blue", "red", "black")
)

head(res,20)

p <- ggplot(res, aes(x=log2FoldChange,y=-log10(pvalue), color=color)) +
  geom_point(size=2.5,alpha=0.85)+
  geom_hline(yintercept=-log10(0.05), linetype="dashed")+
  geom_vline(xintercept=c(-1,1),linetype="dashed")+
  scale_color_identity()+ 
   geom_text(
    data = annot,
    aes(x = x, y = y, label = label, color = color),
    inherit.aes = FALSE,
    size = 5
  )+
  labs(x = "log2FoldChange",
       title=paste0("Volcano Plot of ", exp_name)) +
  theme_classic()
  return(p)
}
```

```{r}
volcanoplots <- list()
for(item in terms){
  volcanoplots[[item]] <- volcano_plot(dds, item)
}
```


```{r}
head(volcanoplots)
```





I want to get a dataframe that has the different groups, and the collective averages, stabilizing sample 9 in one and having a column where it is not stabilized
```{r}
# compute mean expression per group
group_matrix <- sapply(unique(meta$plot_group), function(g) {
  rowMeans(vst[, meta$plot_group == g, drop = FALSE])
})

group_matrix_z <- t(scale(t(group_matrix)))
group_matrix_z[is.na(group_matrix_z)] <- 0

group_data <- list(
  mean = group_matrix,
  zscore = group_matrix_z
)

```

```{r}
top_genes <- apply(group_matrix_z, 2, function(col) {
  names(sort(col, decreasing = TRUE))[1:5]
})

top_genes <- unique(as.vector(top_genes))

heatmap_matrix <- group_data$zscore[top_genes, ]

pheatmap::pheatmap(
  heatmap_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE
)
```

Want to do a heatmap for general study and quick overview for the paper


Limma Array Initializing
```{r}
log_counts <- log2(count_matrix+1)

qn_log_counts <- normalizeBetweenArrays(
     log_counts,
     method = "quantile")
```


We want to do Density Plots for the Raw Counts, DESeq2 VST Counts, and the QN_Log_Counts
```{r}
outliers <- c("Yim-009","Yim-015","Yim-058","Yim-048","Yim-071","Yim-006","Yim-017",
              "Yim-045","Yim-034","Yim-039")

plot_density_with_colored_outliers <- function(mat, title,
                                               log_transform = FALSE,
                                               xlim_zoom = NULL,
                                               ylim_zoom = NULL) {
  
  df <- as.data.frame(mat)
  df$Gene <- rownames(df)
  
  df_long <- df %>%
    pivot_longer(
      cols = -Gene,
      names_to = "Sample",
      values_to = "Value"
    ) %>%
    mutate(
      ColorGroup = ifelse(Sample %in% outliers, Sample, "Other"),
      PlotValue = if (log_transform) log2(Value + 1) else Value
    ) %>%
    arrange(ColorGroup == "Other")  # gray lines first, outliers on top

  p <- ggplot(df_long, aes(x = PlotValue, group = Sample, color = ColorGroup)) +
    geom_density(linewidth = 0.8) +  # lines only
    scale_color_manual(
      values = c(
        "Other" = "grey80",
        setNames(scales::hue_pal()(length(outliers)), outliers)
      )
    ) +
    # filled legend trick
    guides(color = guide_legend(
      override.aes = list(linetype = 0, shape = 22, fill = c("grey80", scales::hue_pal()(length(outliers))))
    )) +
    labs(
      title = title,
      x = if (log_transform) "log2(count + 1)" else "Expression",
      y = "Density",
      color = "Sample"
    ) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold"))
  
  if (!is.null(xlim_zoom) || !is.null(ylim_zoom)) {
    p <- p + coord_cartesian(xlim = xlim_zoom, ylim = ylim_zoom)
  }
  
  return(p)
}




plot_density_with_colored_outliers(
  mat = count_matrix,
  title = "Raw Counts Density",
  log_transform = TRUE,
  xlim_zoom = c(0, 3),
  ylim_zoom = c(0, 0.175)
)

plot_density_with_colored_outliers(
  mat = vst_df,
  title = "VST Density (DESeq2 Variance Stabilized Transform)",
  log_transform = FALSE,
  xlim_zoom = c(4.857424, 7),
  ylim_zoom = c(0, 0.6)
)

plot_density_with_colored_outliers(
  mat = qn_log_counts,
  title = "Quantile-Normalized Log Density",
  log_transform = FALSE,
  xlim_zoom = c(0, 3),
  ylim_zoom = c(0, 0.2)
)


```
Doing density plot for DESeq2 Normalized Counts instaed of VST counts
```{r}
norm_counts_df <- as.data.frame(norm_counts)

print(head(norm_counts_df))

norm_counts_df <- rownames_to_column(norm_counts_df,var="gene")

norm_counts_df <- norm_counts_df %>% pivot_longer(cols=-gene,names_to="sample", values_to="value")

print(head(norm_counts_df))

ggplot(norm_counts_df, aes(x=log1p(value))) +
  geom_density()
```
Looking at dispersoin plot
```{r}
plotDispEsts(dds)
plotSparsity(dds)
```



Computing PCA for DESeq Filtered
```{r}

pca_input <- t(vst)

any(apply(pca_input, 2, var) == 0)


pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  metadata[rownames(pca$x), ]
)

pve <- pca$sdev^2 / sum(pca$sdev^2)
pve[1:5]

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)

pca_df$GroupFactor <- meta[rownames(pca_df), "Group"]
#pca_df$GroupFactor <- factor(pca_df$GroupID)


ggplot(pca_df, aes(x = PC1, y = PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of 8 Groups",
    subtitle="DESeq Filtered Beforehand"
  ) +
  theme_classic()


```




Computing PCA for Quantile Normalization
```{r}
dim(qn_log_counts)              # genes x samples
mode(qn_log_counts)             # should be "numeric"
anyNA(qn_log_counts)            # should be FALSE

gene_var <- apply(qn_log_counts, 1, var)

qn_log_filt <- qn_log_counts[gene_var > 1e-8, ]

pca_input <- t(qn_log_filt)

any(apply(pca_input, 2, var) == 0)


pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  metadata[rownames(pca$x), ]
)

pve <- pca$sdev^2 / sum(pca$sdev^2)
pve[1:5]

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)

pca_df$GroupFactor <- meta[rownames(pca_df), "Group"]
#pca_df$GroupFactor <- factor(pca_df$GroupID)


ggplot(pca_df, aes(x = PC1, y = PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of 8 Groups",
    subtitle="Sample 9 Included along with Ages"
  ) +
  theme_classic()


```

PCA of Quantile with Sample 9 removed
```{r}
keep<- !(
  rownames(meta) == "Yim-009" |
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
qn_sub <- qn_log_counts[, rownames(meta_sub)]

gene_var <- apply(qn_sub, 1, var)
qn_sub_filt <- qn_sub[gene_var > 1e-8, ]

pca_input <- t(qn_sub_filt)

stopifnot(!any(apply(pca_input, 2, var) == 0))

pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of Quantile Normalized Saline Data",
    subtitle="Sample 9 is Removed for Clarity"
  ) +
  theme_classic()
```

Having PCA with sample 9 in it
```{r}
keep<- !(
  meta$Treatment == "rs"
)

meta_sub <- meta[keep, ]
qn_sub <- qn_log_counts[, rownames(meta_sub)]

gene_var <- apply(qn_sub, 1, var)
qn_sub_filt <- qn_sub[gene_var > 1e-8, ]

pca_input <- t(qn_sub_filt)

stopifnot(!any(apply(pca_input, 2, var) == 0))

pca <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  meta_sub[rownames(pca$x), ]
)

pve <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df$GroupFactor <- interaction(
  pca_df$Age,
  pca_df$Exposure,
  pca_df$Diet,
  pca_df$Treatment,
  drop = TRUE
)

pca_df$Outlier <- with(
  pca_df,
  abs(scale(PC1)) > 2 | abs(scale(PC2)) > 2
)


ggplot(pca_df, aes(PC1, PC2, color = GroupFactor)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = subset(pca_df, Outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = paste0("PC1 (", round(pve[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(pve[2] * 100, 1), "%)"),
    color = "Group",
    title="PCA of Quantile Normalized Saline Data",
    subtitle="Sample 9 is is not removed"
  ) +
  theme_classic()

```
Now we will extract these PCA loadings, and get the most important genes from them.
```{r}
loadings <- as.data.frame(pca$rotation[, 1:2])
colnames(loadings) <- c("PC1", "PC2")
loadings$GeneID <- rownames(loadings)

top_pc1 <- loadings |>
  dplyr::arrange(desc(abs(PC1))) |>
  dplyr::slice(1:20)

top_pc2 <- loadings |>
  dplyr::arrange(desc(abs(PC2))) |>
  dplyr::slice(1:20)

top_pc1$Direction <- ifelse(top_pc1$PC1 > 0, "Positive", "Negative")
top_pc2$Direction <- ifelse(top_pc2$PC2 > 0, "Positive", "Negative")

pc1_genes <- top_pc1$GeneID
pc2_genes <- top_pc2$GeneID

#cat(pc1_genes,sep="\n")
#print("DIVIDER\n")
#cat(pc2_genes,sep="\n")

pc1_gene_names <- c("Myh9",
"Rnf11",
"Cltc",
"Col4a1",
"Slc10a1",
"Iqgap1",
"Scrn2",
"Cfap68",
"Hsd11b1",
"Actr2",
"Agfg1",
"Mrps35",
"Sqstm1",
"Fat1",
"Ctnnb1",
"Aamdc",
"Flna",
"Nme3",
"BC024139",
"Ces3b")

pc2_gene_names <- c("Fhit",
"Arhgef16",
"Qprt",
"Hectd1",
"Mfsd12",
"Eno1",
"Dera",
"Acadm",
"Gm70036",
"Nudt7",
"Hsd17b6",
"Cndp2",
"Napa",
"Klk1b4",
"Tmem86b",
"Dnaja3",
"Pgd",
"Frmd8",
"Slc23a1",
"Pcbd1")

top_pc1$GeneName <- pc1_gene_names
top_pc2$GeneName <- pc2_gene_names

top_pc1$Magnitude <- abs(top_pc1$PC1)
top_pc2$Magnitude <- abs(top_pc2$PC2)

top_pc1$GeneName <- reorder(
  top_pc1$GeneName,
  top_pc1$Magnitude
)

top_pc2$GeneName <- reorder(
  top_pc2$GeneName,
  top_pc2$Magnitude
)

ggplot(top_pc1,
       aes(x = GeneName, y = Magnitude, fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "Positive" = "blue",
    "Negative" = "red"
  )) +
  labs(
    x = "Gene",
    y = "PC1 Loading",
    title = "Top Genes Driving PC1"
  ) +
  theme_classic()

ggplot(top_pc2,
       aes(x = GeneName, y = Magnitude, fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "Positive" = "blue",
    "Negative" = "red"
  )) +
  labs(
    x = "Gene",
    y = "PC2 Loading",
    title = "Top Genes Driving PC2"
  ) +
  theme_classic()

```

```{r}
head(meta)
```


Want to do GO enrichment and find top 20 genes of true control 1d vs WSPDsa 1d
```{r}
meta_test <- meta[
  meta$Age %in% c("1 day","28 day") &
  meta$Treatment == "sa" &
  meta$Diet == "PD" &
  meta$Exposure == "WS",
]

table(meta_test$Exposure, meta_test$Age)
meta_test

count_test <- count_matrix[,rownames(meta_test)]

dds <- DESeqDataSetFromMatrix(
  countData = count_test,
  colData   = meta_test,
  design    = ~ plot_group
)

#Potential Filtering Option
dds <- dds[rowSums(counts(dds)) >= 10, ]
dds <- DESeq(dds)

resultsNames(dds)

res <- results(dds)
head(res)
```

```{r}
res <- results(dds, name = "plot_group_WSPDsa.28.day_vs_WSPDsa.1.day")
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)



# Swapping Order so it should be 1 day vs 28 day instead of 28 day vs 1 day
res$log2FoldChange <- -res$log2FoldChange


# keep only significant genes
sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

# add absolute log2 fold change
sig$absLog2FC <- abs(sig$log2FoldChange)

# sort by absolute effect size
sig_sorted <- sig[order(-sig$absLog2FC), ]
sig_sorted$clean_ensembl <- str_remove(sig_sorted$ENSEMBL, "\\.\\d+$")
sig_sorted$SYMBOL <- ids$SYMBOL[match(sig_sorted$clean_ensembl, ids$clean_ensembl)]

# Only get mapped symbols
sig_mapped <- sig_sorted[!is.na(sig_sorted$SYMBOL) & sig_sorted$SYMBOL != "", ]

# look at top hits
head(sig_sorted, 10)

```

```{r}
# take top 20 by absolute log2FC
top20 <- head(sig_mapped, 20)

# clean Ensembl again just in case
top20$clean_ensembl <- str_remove(top20$ENSEMBL, "\\.\\d+$")

# map gene symbols
top20$SYMBOL <- ids$SYMBOL[match(top20$clean_ensembl, ids$clean_ensembl)]

# if some symbols are missing, fall back to Ensembl
top20$LABEL <- ifelse(is.na(top20$SYMBOL), top20$clean_ensembl, top20$SYMBOL)

# classify direction
top20$Direction <- ifelse(top20$log2FoldChange > 0, "Upregulated in WSPDsa 28 day", "Downregulated in WSPDsa 28 day")


ggplot(top20, aes(x = reorder(LABEL, absLog2FC),
                  y = absLog2FC,
                  fill = Direction)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6))+
  labs(
    x = "Gene Name",
    y = "| log2(FoldChange) |",
    title = "Top 20 Log Fold Genes (Mapped Gene Symbols Only)",
    subtitle = "WSPDsa 1 Day vs WSPDsa 28 Day"
  ) +
  theme_classic(base_size = 13)

```

```{r}
sig$clean_ensembl <- str_remove(sig$ENSEMBL, "\\.\\d+$")
sig$ENTREZID <- ids$ENTREZID[match(sig$clean_ensembl, ids$clean_ensembl)]

# remove genes without Entrez IDs
sig <- sig[!is.na(sig$ENTREZID), ]

sig_up   <- sig[sig$log2FoldChange > 0, ]
sig_down <- sig[sig$log2FoldChange < 0, ]

up_entrez   <- unique(sig_up$ENTREZID)
down_entrez <- unique(sig_down$ENTREZID)

# background universe
bg_entrez <- unique(ids$ENTREZID)
bg_entrez <- bg_entrez[!is.na(bg_entrez)]

go_up <- enrichGO(
  gene          = up_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

go_down <- enrichGO(
  gene          = down_entrez,
  universe      = bg_entrez,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)
```

```{r}
go_up_df <- as.data.frame(go_up)
print(head(go_up_df))

# sort by padj ascending
go_up_df <- go_up_df[order(go_up_df$p.adjust), ]

# pick top N categories
topN <- 15
go_up_top <- go_up_df[1:topN, ]


ggplot(go_up_top, aes(x = reorder(Description, -p.adjust), y = Count, fill=p.adjust)) +
  geom_col() +
  scale_fill_gradient(low="#F8766D",high="#619CFF") +
  coord_flip() +
  labs(
    x = "GO Term",
    y = "Gene Count",
    title = "Top 15 GO Terms – Upregulated in WSPDsa 28 Day"
  ) +
  theme_classic(base_size = 13)

go_down_df <- as.data.frame(go_down)
go_down_df <- go_down_df[order(go_down_df$p.adjust), ]
go_down_top <- go_down_df[1:15, ]

print(head(go_down_df))

ggplot(go_down_top, aes(x = reorder(Description, -p.adjust), y = Count, fill=p.adjust)) +
  geom_col() +
  scale_fill_gradient(low="#F8766D",high = "#619CFF")+
  coord_flip() +
  labs(
    x = "GO Term",
    y = "Gene Count",
    title = "Top 15 GO Terms – Downregulated in WSPDsa 28 Day"
  ) +
  theme_classic(base_size = 13)


```


Recreating Jennifers Poster Top 5 Genes Graph
Please Run DDS with all the differing conditions
```{r}
resultsNames(dds)
terms <- setdiff(resultsNames(dds),"Intercept")

terms[1]

res <- results(dds, name = terms[1])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

# List of Genes that we will exclude from the top genes list
exclude <- res[!is.na(res$padj) & res$padj < 0.05, ]

# List of Woodsmoke Genes that are significant
res <- results(dds, name = terms[3])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

# Filtering for only top genes in exposure
exposure_genes <- sig[!sig$ENSEMBL %in% exclude$ENSEMBL, ]

res <- results(dds, name = terms[2])
res <- as.data.frame(res)
res$ENSEMBL <- rownames(res)

sig <- res[!is.na(res$padj) & res$padj < 0.05, ]

common_genes <- intersect(exposure_genes$ENSEMBL,sig$ENSEMBL)
exposure_genes <- exposure_genes[exposure_genes$ENSEMBL %in% common_genes, ]
exposure_genes <- exposure_genes[order(exposure_genes$padj),]

# Now we need to map these to their according gene names
exposure_genes$clean_ensembl <- str_remove(exposure_genes$ENSEMBL, "\\.\\d+$")
exposure_genes$SYMBOL <- ids$SYMBOL[match(exposure_genes$clean_ensembl, ids$clean_ensembl)]


exposure_genes <- exposure_genes[!is.na(exposure_genes$SYMBOL) & exposure_genes$SYMBOL != "", ]
head(exposure_genes$SYMBOL)

# Now we just need to plot all of this with our different groups
expr <- assay(vsd)
rownames(expr) <- str_remove(rownames(expr), "\\.\\d+$")

expr <- expr[ids$clean_ensembl[ids$SYMBOL %in% exposure_genes$SYMBOL[1:5]],,drop=FALSE]
rownames(expr) <- ids$SYMBOL[match(rownames(expr),ids$clean_ensembl)]

expr_long <- as.data.frame(expr) %>%
  mutate(SYMBOL = rownames(.)) %>%
  pivot_longer(
    cols = -SYMBOL,
    names_to = "sample",
    values_to = "expression"
  )

expr_long <- expr_long %>%
  left_join(meta, by = c("sample" = "SampleID"))

meta$plot_group

groups_keep <- c(
  "FAPDsa 1 day",
  "FAWDsa 1 day",
  "WSPDsa 1 day",
  "WSWDsa 1 day",
  "FAPDsa 28 day",
  "WSPDsa 28 day",
  "FAWDsa 28 day",
  "WSWDsa 28 day"
)

expr_long <- expr_long %>%
  filter(plot_group %in% groups_keep)

expr_long$plot_group <- factor(
  expr_long$plot_group,
  levels = groups_keep
)

options(
  repr.plot.width  = 14,
  repr.plot.height = 8
)

ggplot(expr_long, aes(x = plot_group, y = expression, color = plot_group)) +
  stat_summary(
  fun = mean,
  geom = "crossbar",
  width = 0.6,
  color = "black",
  linewidth = 0.5
) +
geom_jitter(
  width = 0.15,
  size = 3,
  alpha = 0.8
) +
  facet_wrap(~ SYMBOL, scales = "free_y") +
  theme_classic(base_size = 13) +
   theme(
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.5),
    panel.grid.minor.y = element_line(color = "grey90", linewidth = 0.25),
    panel.grid.major.x = element_line(color = "grey88", linewidth = 0.4),
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1.09, hjust = 1),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = c(1, 0),      # bottom right
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "black"),  # optional border
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11)
  )  +
  labs(
    x = "Exposure / Age Group",
    y = "VST Normalized Expression",
    color = "Group",
    title = "Top 5 Exposure-Specific Genes"
  )

head(exposure_genes)
```

```{r}

top_genes_all <- lapply(
  setdiff(resultsNames(dds), "Intercept"),
  function(coef_name) {

    res <- results(dds, name = coef_name) %>%
      as.data.frame() %>%
      rownames_to_column("ENSEMBL") %>%
      filter(!is.na(padj), padj < 0.05) %>%
      arrange(desc(abs(log2FoldChange))) %>%
      slice_head(n = 10) %>%
      mutate(term = coef_name)

    res
  }
)

top_genes_all <- bind_rows(top_genes_all)

print(top_genes_all)
top_genes_all$clean_ensembl <- str_remove(top_genes_all$ENSEMBL, "\\.\\d+$")
top_genes_all$SYMBOL <- ids$SYMBOL[match(top_genes_all$clean_ensembl, ids$clean_ensembl)]

top_genes_all <- top_genes_all[!is.na(top_genes_all$SYMBOL) & top_genes_all$SYMBOL != "", ]
print(top_genes_all$SYMBOL)

vst_sub <- vst_df[rownames(vst_df) %in% top_genes_all$ENSEMBL, , drop = FALSE]

rownames(vst_sub) <- top_genes_all$SYMBOL[
  match(rownames(vst_sub), top_genes_all$ENSEMBL)
]

vst_scaled <- t(scale(t(vst_sub)))

vst_long <- vst_sub %>%
  as.data.frame() %>%
  mutate(SYMBOL = rownames(.)) %>%
  pivot_longer(
    cols = -SYMBOL,
    names_to = "SampleID",
    values_to = "expression"
  )

head(meta)
meta$plot_group
# Join metadata (plot_group and groupID)

vst_long <- vst_long %>%
  mutate(SampleID_numeric = as.character(as.integer(str_remove(SampleID, "^Yim-"))))


vst_long <- vst_long %>%
  left_join(dplyr::select(meta, SampleID, plot_group, GroupID), 
            by = c("SampleID_numeric" = "SampleID"))


print(head(vst_long))

avg_df <- vst_long %>%
  group_by(SYMBOL, plot_group) %>%
  summarise(avg_expression = mean(expression), .groups = "drop")

avg_mat <- avg_df %>%
  pivot_wider(names_from = plot_group, values_from = avg_expression) %>%
  column_to_rownames("SYMBOL") %>%
  as.matrix()

avg_scaled <- t(scale(t(avg_mat)))


pheatmap(avg_scaled)


```
Trying out umap for maybe different visualisation techniques
```{r}
vst_umap <- umap(vst)

```

```{r}
#head(vst_umap)

umap_df <- as.data.frame(vst_umap$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

#umap_df$SampleID <- colnames(vst)
head(umap_df)

umap_df$ensembl <- rownames(umap_df)


print(summary(umap_df))

plot(umap_df$UMAP1,umap_df$UMAP2)

# Need to look into classification of genes into potential categories 
# Otherwise cannot look into this one
```

Sample Level UMAp
```{r}
vst_t <- t(vst)

vst_umap_samples <- umap(vst_t, n_components = 3, random_state=52)

```

Analysis
```{r}

head(rownames(vst_umap_samples$layout))
# Example output: "Yim-001", "Yim-002", ...

# Make meta$SampleID match this format
meta$SampleID <- paste0("Yim-", sprintf("%03d", 1:nrow(meta)))

# Now create umap_sample_df
umap_sample_df <- as.data.frame(vst_umap_samples$layout)
colnames(umap_sample_df) <- c("UMAP1", "UMAP2","UMAP3")
umap_sample_df$SampleID <- rownames(umap_sample_df)

# Merge by SampleID
umap_sample_df <- merge(umap_sample_df, meta, by = "SampleID")

head(umap_sample_df)

ggplot(umap_sample_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(size = 3) +
  theme_minimal()
```

```{r}
plot_ly(umap_sample_df, x=~UMAP1,y=~UMAP2,z=~UMAP3,color=~Group,type="scatter3d",mode="markers",marker=list(size=4))
```


Creating function to run through DDS object with every experiment in the dds object and creating a dotplot for each one
```{r}

gsea_enrich <- function(dds, exp_name, ids, top=10){
  res <- results(dds, name=exp_name)
  res <- res[!is.na(res$stat),]
  
  gene_ranks <- res$stat
  names(gene_ranks) <- rownames(res)
  gene_ranks <- sort(gene_ranks,decreasing=TRUE)
  
  gene_df <- data.frame(ensembl_ids = names(gene_ranks),
                        stat = gene_ranks,
                        stringsAsFactors = FALSE)
  gene_df$clean_ensembl <- str_remove(gene_df$ensembl_ids,"\\.\\d+$")

  gene_df$ENTREZID <- ids$ENTREZID[match(gene_df$clean_ensembl, ids$clean_ensembl)]

  gene_df <- gene_df[!is.na(gene_df$ENTREZID) & gene_df$ENTREZID != "",]

  # Order by absolute stat (largest effect first)
  gene_df <- gene_df[order(abs(gene_df$stat), decreasing = TRUE), ]

  # Remove duplicate ENTREZ IDs (keep first = strongest signal)
  gene_df <- gene_df[!duplicated(gene_df$ENTREZID), ]

  gene_ranks <- gene_df$stat
  names(gene_ranks) <- gene_df$ENTREZID

  gene_ranks <- sort(gene_ranks, decreasing = TRUE)
  
  gsea_res <- gseGO(
  geneList = gene_ranks,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "ENTREZID",
  BPPARAM = BiocParallel::SerialParam()
  )
  
  gsea_df <- as.data.frame(gsea_res)
  
  gsea_df <- gsea_df %>% arrange(p.adjust) %>% slice_head(n=top)
  gsea_df$p_label <- paste0("padj = ", signif(gsea_df$p.adjust,3))
  
  gsea_df$Direction <- ifelse(gsea_df$NES > 0,
                            "Upregulated",
                            "Suppressed")
  
  p <- ggplot(gsea_df, aes(x=reorder(Description,NES), y=NES,fill=p.adjust)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient(low="#F8766D",high="#619CFF")+
    labs(
    x = "Pathways/Labels",
    y = "Normalized Enrichment Score (NES)",
    title = paste0("GSEA Enrichment of ", exp_name),
    fill = "Direction"
    ) +
    expand_limits(y = max(abs(gsea_df$NES)) * 1.3) +
    theme_classic()
  
  return(p)
}
```

```{r}
plots <- list()

for (term in terms) {
  plots[[term]] <- gsea_enrich(dds, term, ids)
}
#gsea_enrich(dds,"Age_28.day_vs_1.day",ids)
```

```{r}
head(plots)
```

```{r}
head(meta)
control_group <- "FAPDsa 1 day"
test_group <- "WSPDsa 1 day"

# Helper Function that looks at the control group and test group input
# and is able to extract the group and age parts of the string
parse_group_label <- function(label) {
  parts <- strsplit(label, " ")[[1]]
  list(
    group = parts[1],
    age   = tolower(paste(parts[2], parts[3]))
  )
}


ctrl <- parse_group_label(control_group)
test <- parse_group_label(test_group)

mask_control <-
  meta$Group == ctrl$group &
  meta$Age   == ctrl$age

mask_test <-
  meta$Group == test$group &
  meta$Age   == test$age

keep <- mask_control | mask_test

# Removing Sample 9
#keep[9] <- FALSE

meta_test <- meta[keep,]

# Examining The sample groups
table(meta_test$Age)
meta_test

meta_test$Group <- relevel(
  factor(meta_test$Group),
  ref = ctrl$group
)


count_test <- count_matrix[,rownames(meta_test)]

dds <- DESeqDataSetFromMatrix(
  countData = count_test,
  colData   = meta_test,
  design    = ~ plot_group
)

#Potential Filtering Option
dds <- dds[rowSums(counts(dds)) >= 10, ]
dds <- DESeq(dds)
```

```{r}
term0 <- resultsNames(dds)
term0[2]
plot <- gsea_enrich(dds,term0[2],ids,15)
```

```{r}
plot
```



This is getting the matrix for the counts and then removing sample 9
```{r}
dataExpr <- vst[,-9]
print(colnames(dataExpr))
dataExpr <- t(dataExpr)



traits <- meta %>% select(Exposure,Diet,Treatment,Age)
traits <- traits %>% mutate(across(everything(), ~ as.numeric(as.factor(.))))
traits <- traits[-9,]
print(traits)
```


Now conducting WGCNA to hopefully create a network that we can connect and view for biological intrepretation
```{r}


gene_var <- apply(dataExpr,2,var)
keep <- gene_var > quantile(gene_var,0.75)
dataExpr <- dataExpr[,keep]

gsg <- goodSamplesGenes(dataExpr,verbose=3)
dataExpr <- dataExpr[gsg$goodSamples, gsg$goodGenes]

powers <- 1:20
sft <- pickSoftThreshold(dataExpr,powerVector=powers, verbose = 5)
softPower <- 6

net <- blockwiseModules(
  dataExpr,
  power = softPower,
  TOMType = "signed",
  minModuleSize = 30,
  mergeCutHeight = 0.25,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  verbose = 3
)

moduleColors <- labels2colors(net$colors)

```

```{r}
MEs <- net$MEs
MEs <- orderMEs(MEs)

moduleTraitCor <- cor(MEs, traits, use = "p")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nrow(dataExpr))

# Heatmap of module-trait correlations
labeledHeatmap(
  Matrix = moduleTraitCor,
  xLabels = colnames(traits),
  yLabels = names(MEs),
  colors = blueWhiteRed(50),
  textMatrix = signif(moduleTraitCor, 2),
  main = "Module-trait relationships"
)

# use numeric labels directly
numericMods <- net$colors  # numeric labels
MEs <- net$MEs             # column names like ME1, ME2

hub_genes_list <- list()
hub_scores <- c()

for (modNum in unique(numericMods)) {
  moduleGenes <- numericMods == modNum
  MEname <- paste0("ME", modNum)
  kME <- cor(dataExpr[, moduleGenes], MEs[, MEname], use = "p")
  
  n_top <- ceiling(0.05*sum(moduleGenes))
  top_hubs <- names(sort(kME[,1], decreasing = TRUE))[1:n_top]
  hub_genes_list[[MEname]] <- top_hubs
  hub_scores <- c(hub_scores, kME[top_hubs, 1])
}
# flatten hub genes from all modules
top_genes <- unique(unlist(hub_genes_list))
hub_scores <- hub_scores[match(top_genes, names(hub_scores))]
# adjacency matrix
cor_mat <- cor(dataExpr[, top_genes], use="p")  # Pearson correlation
adj_mat <- abs(cor_mat)^softPower  # weighted adjacency

# optionally threshold adjacency for plotting clarity
adj_mat[adj_mat < 0.2] <- 0

pheatmap(adj_mat, cluster_rows = TRUE, cluster_cols = TRUE)
```

```{r}
g <- graph_from_adjacency_matrix(adj_mat, mode = "undirected", weighted = TRUE, diag = FALSE)

# Color nodes by module
V(g)$color <- moduleColors[match(V(g)$name, colnames(dataExpr))]

# Scale node size by hub score
V(g)$size <- 5 + 10 * hub_scores[match(V(g)$name, top_genes)]

# Remove labels
V(g)$label <- NA

dev.new(width = 18, height = 16)

png("hub_gene_network.png", width = 3000, height = 2500, res = 300)

lay <- layout_with_fr(g)
lay <- lay * 2   # increase spacing

plot(g,
     layout = lay,
     vertex.size = V(g)$size,
     vertex.color = V(g)$color,
     edge.width = E(g)$weight * 2,
     main = "Top Hub Gene Network"
)
```

```{r}
gene_module_df <- data.frame(
  ensembl_ids = colnames(dataExpr),
  Module = moduleColors
)

modules <- unique(moduleColors)

module_gene_list <- lapply(modules, function(mod){
  gene_module_df$Gene[gene_module_df$Module == mod]
})

names(module_gene_list) <- modules

# Write this result to a csv so we can load it later without running everything again
wb <- createWorkbook()
modules <- unique(gene_module_df$Module)

for (mod in modules) {

  module_genes <- gene_module_df[
    gene_module_df$Module == mod, ]

  addWorksheet(wb, mod)

  writeData(wb,
            sheet = mod,
            module_genes)
}

saveWorkbook(wb,
             "WGCNA_modules.xlsx",
             overwrite = TRUE)
```

```{r}
head(gene_module_df)

gene_module_df <- gene_module_df %>% left_join(ids,by="ensembl_ids")
head(gene_module_df)
print(length(gene_module_df$ensembl_ids))
```

Function that goes through the WGCNA module color provided and the gene table, and does the go enrichment for all genes that have that color module
```{r}
run_module_GO <- function(df, module_color,
                          ont = "BP",
                          p_cutoff = 0.05) {

  # 1️⃣ Filter module + valid Entrez IDs
  entrez_ids <- df %>%
  filter(Module == module_color) %>%
  filter(!is.na(ENTREZID)) %>%
  pull(ENTREZID) %>%
  unlist() %>%        # ⭐ CRITICAL FIX
  as.character() %>%  # ensure atomic vector
  unique()

  # Safety check
  if (length(entrez_ids) == 0) {
    stop(paste("No valid ENTREZIDs for module:", module_color))
  }

  # 2️⃣ Background universe (all measured genes)
  background <- df %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unlist() %>%
    as.character() %>%
    unique()

  # 3️⃣ GO enrichment
  ego <- enrichGO(
    gene          = entrez_ids,
    universe      = background,
    OrgDb         = org.Mm.eg.db,
    keyType       = "ENTREZID",
    ont           = ont,
    pAdjustMethod = "BH",
    pvalueCutoff  = p_cutoff,
    readable      = TRUE
  )

  return(ego)
}
```


Creating a function that will run through every module and download the png of the go enrichment
```{r}
plot_module_GO <- function(go_df, module_color, top_n = 15, outdir = "Graphs") {
  if(is.null(go_df) || nrow(go_df) == 0){
    message(paste("No data to plot for module:", module_color))
    return(NULL)
  }
  
  # Select top terms by adjusted p-value
  top_terms <- go_df %>%
    arrange(p.adjust) %>%
    slice_head(n = top_n)
  
  top_terms$Description <- factor(top_terms$Description, levels = top_terms$Description)
  
  # Plot
  p <- ggplot(top_terms,
              aes(x = reorder(Description, Count),
                  y = Count,
                  fill = p.adjust)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient(low = "red", high = "blue") +
    labs(
      x = "",
      y = "Gene Count",
      title = paste("GO Enrichment - Module", module_color)
    ) +
    theme_minimal(base_size = 14)
  
  # Save plot
  if(!dir.exists(outdir)) dir.create(outdir)
  ggsave(filename = file.path(outdir, paste0("GO_barplot_", module_color, ".png")),
         plot = p, width = 16, height = 14)
  
  return(p)
}
```

Loop running through every color and saving results as png in graphs
```{r}
for(color in unique(moduleColors)){
  go <- as.data.frame(run_module_GO(gene_module_df, color))
  plot_module_GO(go,color)
}
```


```{r}

# Subset your original meta dataframe to only include these samples
# Get the samples actually in your expression matrix
network_samples <- rownames(dataExpr)

# Subset meta by these rownames
network_meta <- meta[network_samples, , drop = FALSE]  # keeps rownames

# Create numeric trait columns
network_meta$Exposure_num  <- ifelse(network_meta$Exposure == "WS", 1, 0)
network_meta$Diet_num      <- ifelse(network_meta$Diet == "PD", 1, 0)
network_meta$Treatment_num <- ifelse(network_meta$Treatment == "rs", 1, 0)
network_meta$Age_num <- ifelse(network_meta$Age == "1 day", 1,0)
# Select only numeric traits
traitData <- network_meta[, c("Exposure_num", "Diet_num", "Treatment_num","Age_num")]

# Now correlate with modules
moduleTraitCor <- cor(MEs, traitData, use = "pairwise.complete.obs")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nSamples = nrow(dataExpr))

# Plot heatmap
pheatmap(
  moduleTraitCor,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  main = "Module-Trait Correlations"
)
```





To DO List:
Clean up the network graph and extract biological meaning by either putting pathways or actual gene names on the graph
Do correlation graphs, with gene-gene correlations and gene-trait correlations
Look into the sex differences of mice and see if there is any female specific genes we to exclude from the study, or be filtered in some other parts of the analyis so it does not show up implicity in different parts of the analysis like in the GO enrichment.

















